Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-12-04 paint-contour-tree.tex
critical join replaced with join

\InNotSoCGVer{
 \newtheorem{theorem}{Theorem}[section]
}
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}[theorem]{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
\InNotSoCGVer{
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{example}[theorem]{Example}
 \newtheorem{definition}[theorem]{Definition}
}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
\InNotSoCGVer{ 
 \newtheorem{remark}[theorem]{Remark} 
}
\newtheorem{observation}[theorem]{Observation}

@@

\begin{remark}
The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
it suffices to have an efficient algorithm for computing $\jc(\MM)$.
Due to minor technicalities, the following section is instead phrased in terms of computing $\cJ(\MM)$ efficiently.  
Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
\end{remark}

@@

\smallskip
\begin{compactenum}
	\item Run $\init(\MM)$.
	\item While there are unprocessed critical points:
	\begin{compactenum}
		\item Run $\update(\stack)$. Pop $\stack$ to get $h$.
		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
		\item For all $c' \in \cur(h)$:
		\begin{compactenum}
			\item Add edge $(\h(c'),h)$ to $\jc(\MM)$.
			\item Delete $h$ from $T(c')$.
		\end{compactenum}
	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
	\item Take union of $\cur(h)$ and denote resulting color by $\widehat{c}$.
	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

\begin{claim} \label{clm:process} The partial output on the processed vertices is exactly
the restriction of $\jc(\MM)$ to these vertices.
\end{claim}

\begin{proof} More generally, we prove the following: all outputs on processed vertices
are edges of $\jc(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
of that color. We prove this by induction on the processing order. The base case is trivially
true, as initially the processed vertices and attachments of the color classes are the set of maxima.
For the induction step, consider the situation
when $v$ is being processed.

Since $v$ is being processed, we know by \Cor{terminate} that it is ripe. Take any up-star of $v$, and the corresponding component $\PP$
of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
have the same color (say $c$). If some critical vertex in $\PP$ is not processed,
it must be in $T(c)$, which violates the ripeness of $v$.
Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\jc(\MM)$ to $\PP$ has been correctly computed.
Additionally, since all critical vertices in $\PP$ have processed, they all have the same color $c$ of the lowest critical vertex in $\PP$.
Thus by the strengthened induction hypothesis, this lowest critical vertex is $\h(c)$.

If there is another component of $\MM^+_v$, the same argument implies 
the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
Now by the definition of $\jc(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).

@@

\begin{theorem}
\label{thm:correct}
 Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\jc(\MM)$.
\end{theorem}
\begin{proof}
 First observe that each vertex can be processed at most once by $\build(\MM)$.  By \Cor{terminate}, we know that as long as there 
 is an unprocessed vertex, $\update(K)$ will be called and will terminate with a ripe vertex which is ready to be processed. 
 Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
 computed $\jc(\MM)$.
\end{proof}

@@

\begin{definition}
\label{def:initialColoring}
Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
according to the priority queue for each up-star we put a given vertex in.  Call this the \emph{initial coloring}
of the vertices in $\jc(\MM)$.  Note also that this initial coloring defines the $H_v$ values for all $v\in \jc(\MM)$.
\end{definition}

@@

\begin{lemma}
\label{lem:runTimeUpper}
Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\jc(\MM)$, 
let $H_v$ be defined by the initial coloring of $\MM$.
The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
\end{lemma}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\proofofRunTime}{
\begin{proof}
First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(N)$ time.
Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e.\ each edge is viewed at most twice).
Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.

@@

What remains is the time it takes to process a vertex $v$ in $\build(\MM)$.  
In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
Therefore the only thing left to bound are the queue operations.
Let $v$ be a vertex in $\jc(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
At the time $v$ is processed, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
In particular, when $v$ is processed we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
these merged queues are the queues of the current colors of $c_1$ and $c_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
the only vertices left in these queues are those in $H_{c_1}$ and $H_{c_2}$.

@@

The algorithm $\build(\MM)$ processes vertices in $\jc(\MM)$ one at a time.  The main processing cost  
comes from priority queue operations.  The cost of these operations is a function of the size of the queue which is in turn a function of the choices made by the subroutine $\init(\MM)$.

@@

\begin{definition}\label{def:initialColoring}
A \emph{leaf assignment} $\chi$ of a binary tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
one from the left and one from the right subtree of $v$ (or only one leaf if $v$ has only one child). % naturally if $v$ has only one child it is assigned only one color). 
The subroutine $\init(\MM)$ naturally defines a leaf assignment to $\jc(\MM)$ (which 
is a rooted binary tree) according to the priority queue for each up-star we put a given vertex in.  
Call this the \emph{initial coloring} of the vertices in $\jc(\MM)$, and denote it by $\chi$.

For a vertex $v$ in $\jc(\MM)$, let $L(v)$ denote the set of leaves of the subtree rooted at $v$, and let $A(v)$ denote the set of 
ancestors of $v$, i.e.\ the vertices on the $v$ to root path. For a vertex $v\in \jc(\MM)$, and an initial coloring $\chi$, 
we use $H_v$ to denote the \emph{heap} at $v$. Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(v) \neq \emptyset\}$, i.e.\
the set of ancestors colored by some leaf in $L(v)$.
\end{definition}

@@

\begin{lemma}\label{lem:runTimeUpper}
Let $\MM$ be a simplicial complex with $t$ critical points.  
The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
%For every vertex in $\jc(\MM)$, let $H_v$ be defined by the initial coloring of $\MM$.
%The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
\end{lemma}

@@

% While the above gives a full running time characterization, ideally the bound should be expressed as   
% a function of some structural property of $\jc(\MM)$.
% This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
% the $H_v$ sets (defined by any leaf assignment) naturally induces a path decomposition of $T$ (see \Def{path}), 
% denoted $\pmax(T)$ (see \Sec{pathDecomp} for details).   
% 
% % Specifically, the main result of \Sec{pathDecomp} is the following theorem, which together with the above 
% % bound in terms of heap size immediately implies our main result, Theorem~\ref{thm:main-alg} .
% 
% \begin{theorem} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
% \end{theorem}

% The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
% However, for the specific when $T=\jc(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
% to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
% 
% The path decomposition $\pmax(T)$ is a function of the choices made by the linear time $\init(\MM)$.  
% One cannot control which path decomposition $\init$ determines, as sorting can be reduced to finding the optimal path decomposition.  
% Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
% In the following a rain consistent path decomposition is any decomposition which is the union of paths from 
% decompositions on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).

@@

In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
(this assumption simplifies the presentation but is not necessary).

@@

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
\end{proof}

@@

\begin{theorem}
%Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$.
Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the join tree $\jc(\MM)$ has maximum degree $3$.
There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$.
\end{theorem}
\begin{proof}
By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\jc(\MM)$, and by \Lem{runTimeUpper}
we know this takes $O(\sum_{v \in \jc(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$ 
time to correctly compute $\jc(\MM)$.
\end{proof}

@@

Now one of the key observations was that for extremum dominant manifolds, computing the contour tree is roughly the same as 
computing the join tree.  Specifically, \Thm{contour-tree} implies that given $\jc(\MM_i)$ , 
we can obtain $\cC(\MM_i)$ by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
However, by the above theorem we know $\jc(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
(where $t_i$ and $N_i$ are the number of critical points and faces when restricted to $\MM_i$).

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\jc(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i} = O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]

@@

\begin{remark}
 Note that for the terrains described in this section, the number of critical points is within a constant factor of 
 the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
\end{remark}
}

>>>>

\InNotSoCGVer{
 \newtheorem{theorem}{Theorem}[section]
}
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}[theorem]{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
\InNotSoCGVer{
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{example}[theorem]{Example}
 \newtheorem{definition}[theorem]{Definition}
}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
% \InNotSoCGVer{ 
%  \newtheorem{remark}[theorem]{Remark} 
% }
\newtheorem{Remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

@@

\begin{Remark}\label{rem:joinAndCriticalJoin}
The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
it suffices to have an efficient algorithm for computing $\jc(\MM)$.
Due to minor technicalities, it is easier to phrase the following section instead in terms of computing $\cJ(\MM)$ efficiently.  
Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
\end{Remark}

@@

\smallskip
\begin{compactenum}
	\item Run $\init(\MM)$.
	\item While there are unprocessed critical points:
	\begin{compactenum}
		\item Run $\update(\stack)$. Pop $\stack$ to get $h$.
		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
		\item For all $c' \in \cur(h)$:
		\begin{compactenum}
			\item Add edge $(\h(c'),h)$ to $\cJ(\MM)$.
			\item Delete $h$ from $T(c')$.
		\end{compactenum}
	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
	\item Take union of $\cur(h)$ and denote resulting color by $\widehat{c}$.
	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

\begin{claim} \label{clm:process} The partial output on the processed vertices is exactly
the restriction of $\cJ(\MM)$ to these vertices.
\end{claim}

\begin{proof} More generally, we prove the following: all outputs on processed vertices
are edges of $\cJ(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
of that color. We prove this by induction on the processing order. The base case is trivially
true, as initially the processed vertices and attachments of the color classes are the set of maxima.
For the induction step, consider the situation
when $v$ is being processed.

Since $v$ is being processed, we know by \Cor{terminate} that it is ripe. Take any up-star of $v$, and the corresponding component $\PP$
of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
have the same color (say $c$). If some critical vertex in $\PP$ is not processed,
it must be in $T(c)$, which violates the ripeness of $v$.
Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\cJ(\MM)$ to $\PP$ has been correctly computed.
Additionally, since all critical vertices in $\PP$ have processed, they all have the same color $c$ of the lowest critical vertex in $\PP$.
Thus by the strengthened induction hypothesis, this lowest critical vertex is $\h(c)$.

If there is another component of $\MM^+_v$, the same argument implies 
the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
Now by the definition of $\cJ(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).

@@

\begin{theorem}
\label{thm:correct}
 Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\cJ(\MM)$.
\end{theorem}
\begin{proof}
 First observe that each vertex can be processed at most once by $\build(\MM)$.  By \Cor{terminate}, we know that as long as there 
 is an unprocessed vertex, $\update(K)$ will be called and will terminate with a ripe vertex which is ready to be processed. 
 Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
 computed $\cJ(\MM)$.
\end{proof}

@@

\begin{definition}
\label{def:initialColoring}
Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\cJ(\MM)$ 
according to the priority queue for each up-star we put a given vertex in.  Call this the \emph{initial coloring}
of the vertices in $\cJ(\MM)$.  Note also that this initial coloring defines the $H_v$ values for all $v\in \cJ(\MM)$.
\end{definition}

@@

\begin{lemma}
\label{lem:runTimeUpper}
Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\cJ(\MM)$, 
let $H_v$ be defined by the initial coloring of $\MM$.
The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
\end{lemma}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\proofofRunTime}{
\begin{proof}
First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(N)$ time.
Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e.\ each edge is viewed at most twice).
Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.

@@

What remains is the time it takes to process a vertex $v$ in $\build(\MM)$.  
In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
Therefore the only thing left to bound are the queue operations.
Let $v$ be a vertex in $\cJ(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
At the time $v$ is processed, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
In particular, when $v$ is processed we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
these merged queues are the queues of the current colors of $c_1$ and $c_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
the only vertices left in these queues are those in $H_{c_1}$ and $H_{c_2}$.

@@

The algorithm $\build(\MM)$ processes vertices in $\cJ(\MM)$ one at a time.  The main processing cost  
comes from priority queue operations.  The cost of these operations is a function of the size of the queue which is in turn a function of the choices made by the subroutine $\init(\MM)$.

@@

\begin{definition}\label{def:initialColoring}
A \emph{leaf assignment} $\chi$ of a binary tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
one from the left and one from the right subtree of $v$ (or only one leaf if $v$ has only one child). % naturally if $v$ has only one child it is assigned only one color). 
The subroutine $\init(\MM)$ naturally defines a leaf assignment to $\cJ(\MM)$ (which 
is a rooted binary tree) according to the priority queue for each up-star we put a given vertex in.  
Call this the \emph{initial coloring} of the vertices in $\cJ(\MM)$, and denote it by $\chi$.

For a vertex $v$ in $\cJ(\MM)$, let $L(v)$ denote the set of leaves of the subtree rooted at $v$, and let $A(v)$ denote the set of 
ancestors of $v$, i.e.\ the vertices on the $v$ to root path. For a vertex $v\in \cJ(\MM)$, and an initial coloring $\chi$, 
we use $H_v$ to denote the \emph{heap} at $v$. Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(v) \neq \emptyset\}$, i.e.\
the set of ancestors colored by some leaf in $L(v)$.
\end{definition}

@@

\begin{lemma}\label{lem:runTimeUpper}
Let $\MM$ be a simplicial complex with $t$ critical points.  
The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
%For every vertex in $\cJ(\MM)$, let $H_v$ be defined by the initial coloring of $\MM$.
%The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
\end{lemma}

@@

% While the above gives a full running time characterization, ideally the bound should be expressed as   
% a function of some structural property of $\cJ(\MM)$.
% This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
% the $H_v$ sets (defined by any leaf assignment) naturally induces a path decomposition of $T$ (see \Def{path}), 
% denoted $\pmax(T)$ (see \Sec{pathDecomp} for details).   
% 
% % Specifically, the main result of \Sec{pathDecomp} is the following theorem, which together with the above 
% % bound in terms of heap size immediately implies our main result, Theorem~\ref{thm:main-alg} .
% 
% \begin{theorem} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
% \end{theorem}

% The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
% However, for the specific when $T=\cJ(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
% to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
% 
% The path decomposition $\pmax(T)$ is a function of the choices made by the linear time $\init(\MM)$.  
% One cannot control which path decomposition $\init$ determines, as sorting can be reduced to finding the optimal path decomposition.  
% Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
% In the following a rain consistent path decomposition is any decomposition which is the union of paths from 
% decompositions on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).

@@

In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \Def{criticalJoin}).
We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
(this assumption simplifies the presentation but is not necessary).

@@

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
\end{proof}

@@

\begin{theorem}
%Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\cJ(\MM)$ has maximum degree $3$.
Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the join tree $\cJ(\MM)$ has maximum degree $3$.
There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$.
\end{theorem}
\begin{proof}
By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\cJ(\MM)$, and by \Lem{runTimeUpper}
we know this takes $O(\sum_{v \in \cJ(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$ 
time to correctly compute $\cJ(\MM)$.
\end{proof}

@@

Now one of the key observations was that for extremum dominant manifolds, computing the contour tree is roughly the same as 
computing the join tree.  Specifically, \Thm{contour-tree} implies that given $\jc(\MM_i)$ , 
we can obtain $\cC(\MM_i)$ by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
Remark~\ref{rem:joinAndCriticalJoin} implies that $\jc(\MM_i)$ is trivially obtained from the $\cJ(\MM_i)$, and 
 by the above theorem we know $\cJ(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
(where $t_i$ and $N_i$ are the number of critical points and faces when restricted to $\MM_i$).

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\jc(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\InSoCGVer{
\begin{align*}
\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}&
\end{align*}
}
\InNotSoCGVer{
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]
}

@@

\begin{Remark}
 Note that for the terrains described in this section, the number of critical points is within a constant factor of 
 the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
\end{Remark}
}
